Increasing prevalence of plant-fungal symbiosis over 195 years of climate change

### 2023-03-22 Manuscript/endo_herbarium.tex
edits to manuscript and addding more plots, still working on discussion

	\section*{Abstract}
Climate change has altered species’ phenologies, has triggered changes in traits, and has led to shifts in abundance and distributions. 
Yet, we still know little about how biotic interactions alter these responses over long time scales across spatially heterogenous changes in climate. 
Most species host beneficial microbial symbionts, \tom{which may provide resilience to changes in the environment, but at the same time increasing stress may lead these symbioses to breakdown}{A little confusing as written. Might be better to present as alernative hypotheses}. 
Microbes can be preserved within the dried plant tissue of herbarium specimens presenting an opportunity to quantify changes across broad temporal and spatial scales. 
\tom{Epichloë fungal endophytes are seed-transmitted symbionts of cool-season grasses which can provide drought tolerance to their hosts. }{I would cut this sentence and incorporate need to know info in other sentences.}
\tom{ We ask how endophyte prevalence has responded to climate change.}{Could be sharpened.} 
 Specifically, we took samples of seeds from herbarium specimens of three grass host species collected over the last two centuries, quantified the presence or absence of fungal endophytes within the seeds with microscopy, and evaluated spatial and temporal trends in prevalence using \tom{spatial modeling}{Vague}.
 We found strong evidence that fungal endophyte prevalence has increased over the last two centuries across the three host species from 25\% prevalence to 75\% prevalence on average.
\tom{We also found that observed changed in annual  and seasonal climate drivers were associated with temporal trends in endophyte prevalence. 
In particular, changes in prevalence within each species corresponded to climate drivers during its peak growing season. }{Could be combined into one sentence.}
Thus we provide novel evidence for \tom{adaptive biotic responses}{Not sure what you mean by this.} to ongoing climate change.

@@

	% The journal does not have numbered sections in the main portion of
	% articles. Please refrain from using section references (à la
	% section~\ref{section:CountingOwlEggs}), and refer to sections by name
	% (e.g. section ``Counting Owl Eggs'').
Predicting how species' interactions will be altered by ongoing climate change is a major goal of basic ecological research and conservation management. 
\tom{Spatially heterogeneous changes in the environment have led species to move, adapt, or face extinction \citep{IPCC2012managing}. }{In theory a spatially homogenous change in temperature -- everywhere increases by 2 degrees -- would have the same effect. So significance of spatial heterogeneity here is not clear.}
Species' responses are likely mediated by the numerous interspecific interactions in which they engage. 
Yet for most species, empirically evaluating the role that biotic interactions can play in facilitating or hindering \tom{adaption}{I am struck by this word again. Are you intentionally bringing in evolutionary ideas?} to changing environmental conditions has been limited by a lack of data at adequate spatial and temporal scales \cite{gilman2010framework}.\tom{}{This paragraph has introduced two related but different ideas: how species interactions per se respond to climate change, and how interactinos mediate responses to climate change. This is a little disoroienting.}

Natural history collections, which were originally collected to study and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in ecological interactions across broad spatial scales \citep{meineke2018unrealized}. 
Natural history collections, built and maintained by the efforts of thousands of scientists, are \tom{invaluable time machines}{Love this!}. 
These collections primarily comprise physical specimens of organisms  along with information about the time and place of their collection. 
These specimens are samples preserving physical legacies of ecological dynamics and species' interactions from dynamically changing environments across time and space.
\tom{Recent works using plant collections have documented changes in phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, as well as pollination \citep{pauw2011reconstruction, duan2019century}, and changes in antagonistic interactions such as herbivory \citep{meineke2019herbarium}, but to date, focus has been lacking on long-term changes in a particularly common type of biotic interaction, microbial symbioses.}{I like the content of this sentnece but  found the phrasing difficult.}

Microbial symbionts are common to all macroscopic organisms and can have important effects on their hosts' survival, growth and reproduction \cite{rodriguez2009fungal,mcfall2013animals}.
Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts and often ameliorating host responses to environmental stress. 
For example, bacterial symbionts of insects, such as \emph{Wolbachia}, can improve their hosts' thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}, and arbuscular mycorrhizal fungi, documented in 70-90\% of families of land plants \citep{parniske2008arbuscular}, allow their hosts to persist through drought conditions by improving water and nutrient uptake \citep{cheng2021elucidating}.
On the other hand, changes in the mean and variance of environmental conditions may disrupt microbial  mutualisms by changing the costs and benefits of the interaction for each partner, leading to breakdown of the interaction \citep{aslan2013mutualism}. 
Coral bleaching (the loss of symbiotic algae) due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the community composition of their algal symbionts \citep{meyer2022climate}.
How commonly and under what conditions microbial mutualisms breakdown or strengthem under climate change remain unanswered questions, but previous work suggests that these alternative responses may depend on the intimacy and specialization of the interaction as well as the physiological tolerances of the mutualists \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}. 
\tom{Given that climate change is spatially heterogenous, the rate of change may have a strong influence on when and where we can expect facilitation or disruption of mutualism. }{This is an important sentence that needs some elboration and citation. Can you explain more about spatial heterogeneity in climate change? And are you suggesting that the rate of change itself determines whether mutualism strengthens or breaks down? Also, ``facilitation'' is awkward in this context.}

Early research on \emph{Epichloë} fungal endophytes, specialized symbionts of cool-season grasses, \tom{used herbarium specimens}{I would move this later in the paragraph, where you can make the point that this approach has not been leveraged to its full capacity.} to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}. 
These facultative symbionts, which have been documented in $\sim 30$\% of cool-season grasses \citep{leuchtmann1992systematics}, are transmitted vertically from maternal plants to offspring through the grasses' seeds.
Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
Over time, endophytes that act as mutualists for their hosts should rise in prevalence within a population. 
\emph{Epichloë} are known to improve \tom{drought}{not just drought - also herbivory and pathogens. I would cite more to support the types of endo benefits.} tolerance for their hosts \citep{afkhami2014mutualist}, and are expected to reach high prevalence where these fitness benefits are most important.
Previous survey studies have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
We predicted that \tom{prevalence should also track temporal changes in environmental drivers}{explain and elaborate}, but testing this has been impossible with surveys limited to a few years, and to the \tom{spatial extent reachable during a single growing season}{we have cars and planes so this should not be an obstacle. I think the real limitation is the combination spatio-temporal coverage, which is what herbaria uniquely provide}. 
Becauses the fungi are seed-transmitted, they are conveniently preserved \tom{along with their host plants on herbarium sheets}{as long as the specimen has seeds, which is almost always the case for grasses}, creating the opportunity to examine the response of the symbiosis to historical climate change.

In this study, we took advantage of the spatial and temporal sampling within herbaria to assess for the first time whether grass-fungal endophyte symbioses have responded \tom{adaptively}{the adaptive nature of the response needs to be explained} to climate change. In particular, we (i) quantified the spatial and temporal change in endophyte prevalence across host ranges over the last two centuries, and (ii) identified climate drivers that can explain the magnitude of these responses. \tom{}{This paragraph would benefit from a specific hypothesis or question, which I think you have. Your emphasis on space vs time is a little muddled. You frame the paper around climate change, but part of your analysis - I think - is asking a purely spatial question. This paragraph should also mention the three host species and provide some wow-factor summary stats on the range of years, locations, and sample sizes.}

\section*{Methods}
        \subsection*{Focal host species}
We focused on three endophyte hosts, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.\tom{}{Describe the fungal species.}
These cool-season (C3) grass species are commonly represented in natural history collections with broad distributions covering the Eastern United States.
\emph{A. hyemalis} is a small short-lived perennial species that germinates in the spring and typically blooms between March and July.
\emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and blooms in late Summer and early Autumn. 
This species is relatively more  sparsely distributed, tending to be found in shadier and more moist habitats while \emph{A. hyemalis} is often found in open and recently disturbed ground. 
Both\emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A. perennans} has a slighty more northern distribution, whereas \emph{A. hyemalis} is found rarely as far north as Canada and is listed as a rare plant in Minnesota.
\emph{E. virginicus} is a larger and relatively longer-lived  species that is more broadly distributed that the \emph{Agrostis} species. 
It begins flowering as early as March or April but continues throughout the Summer.

		\subsection*{Herbarium surveys}
We collected 5-10 seeds per specimen during visits to natural history collections between 2019 and 2022 \tom{(See Table A1 for a summary of collections)}{It is important to say more in the text here about where you went and what you did.} . 
\tom{Grasses are commonly identified and collected based on the presence of their reproductive structures which ensured that specimens we examined contained seeds.}{Move to intro} 
\tom{Our sampling plan was designed to minimize damage to these specimens.}{How so?}

\begin{figure}[H]
	\centering
	\includegraphics[width = \linewidth]{collections_map.png}
	\caption{\textbf{Collection locations of herbarium specimens sampled for endopnhyte presence absence}. Color designates host species ( \emph{A. hyemalis} (green), \emph{A. perennans} (orange), \emph{E. virginicus} (purple))}
\end{figure}

After collecting seeds, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, using microscopy. 
We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue dye and squashed them under a microscope cover slip. 
We examined the squashed seeds for the presence of fungal hyphae at 100X magnification \cite{bacon2018stains}.
In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded from the seed counts for that specimen.
On average we scored \textbf{number} seeds per specimen.
Due to imperfect vertical transmission (cite), it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
We therefore designated each specimen as symbiotic if endophyte hyphae were observed in one or more seeds, or non-symbiotic if hyphae were observed in zero seeds. 
To capture uncertainty in the identification process, we recorded both a "liberal" and a "conservative" endophyte status for each plant.  
When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be a true endophyte) and a negative conservative status (less likely to be a true endophyte). 
$89$\% of scored plants had matching liberal and conservative endophyte statuses, reflecting high confidence in endophyte identification.
\tom{The following analyses presented in the main text used the liberal statuses, but we repeated all analyses with the conservative statuses which yielded qualitatively similar results (Appendix). }{Good}

We excluded specimens for which information about the collection location and date were unavailable.
In total, we quantified endophyte symbiosis for $971$ \emph{A. hyemalis}, $302$ \emph{A. perennans}, and $605$ \emph{E. virginicus} specimens spanning the years \tom{$1824$ to $2019$}{give date range by species}.
\tom{Each specimen was assigned a collection location based on geographic information recorded at the time of collection using the ggmap package in R \citep{kahle2019package}.}{confusing-not sure what this means}
Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
For a few of the oldest specimens, only information at the state level was available, and so we used the state centoid \tom{(Fig. 1; Fig A1).}{I don't see how this is shown in figure 1. Also, can you use more distinctive colors in Fig 1?}

\subsection*{Assessing spatial and temporal changes in endophyte prevalence}
\tom{}{Give a quick overview of the types of models you used and what specific purpose they were built for before getting into details. The fact that you estimated climate effects following an initial climate-implicit model needs to come through more clearly.}To quantify spatial and temporal trends in endophyte prevalence, we modeled endophyte presence/absence ($P$) as a Bernoulli response variable for specimen $i$ of host species $h$.

@@

The expected endophyte prevalence, $\hat{P}$, was modelled with intercepts specific to each host species ($\beta_{0}$), slopes for changes over time ($\beta_{1}$) \tom{as well as the interaction between time and the specimen's latitude and longitude ($\beta_{2}$, $\beta_{3}$, and $\beta_{4}$). }{So you do not have equal emphasis on space and time. YOu are really asking a time question and considering how temporal trends may vary spatially. I think this needs to come through more clearly in the introduction.}
We accounted for potential biases introduced during the process of collecting specimens as well as in scoring ability by including random effects specific to each collector $\chi$ and to each scorer $\omega$.
To quantify spatial patterns in prevalence, we included a spatially-dependent random intercept, $\phi$, and fit the model using integrated nested Laplace approximation, an approximate Bayesian method implemented in the rINLA package \citep{lindgren2015bayesian}. 
This method allowed us to model spatial dependency between data points (nearer collections may be more similar) as a continuous process with stochastic partial differential equations (SPDE) that depend on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}. 
The covariance matrix is approximated using a Matérn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).
We fit the model with vague priors, and compared models with different sizes of mesh, which had little effect on the resulting model estimates.
\tom{We assessed model fit with graphical postterior predictive checks (Fig. A3, Fig. A4).}{This is good but also worth mentioning that you assessed convergence.}

\tom{We used data on endophyte prevalence from  contemporary surveys of \emph{A. hyemalis} as test data to evaluate model fit. }{This is an extraordinary extra step that should be played up in the Intro and included in the aims at the end of the Intro.}
During these surveys, which took place between 2015 and 2020, we collected seeds from living plants from 43 populations across a portion of the species' range and then quantified endophyte status with staining microscopy as described for the herbarium surveys (Fig S2).
We compared the model's prediction for these locations to the endophyte status of randomly sampled plants from each surveyed location, \tom{mimicing the process of collecting a single herbarium specimen from the population.}{But isn't the prevalence the better test of the model's predictions?}
The model performed adequately (\tom{training data: AUC = 0.77; test data: AUC = 081}{Is this 0.81? The model performs better on the test data than the training data? This is odd.}; Fig. A4).

		\subsection*{Assessing the role of climate drivers}
To assess how the rate of climate change may have driven changes in endophyte prevalence,\tom{ we calculated the change in climate between recent (1990 to 2020) and historic (1895 to 1925) periods.}{This is where your hypotheses and predictions in the intro need to come through more clearly. Also, this assumes change has been constant during those periods and does not use most years of data. Could you try estimating a slope for each location? Is this possible?}
We first downloaded monthly temperature and precipitation rasters from the PRISM climate group \citep{daly2013prism} covering the time period between 1895 and 2020 using the 'prism' R package \citep{Rprism2015}. 
Prism provides reconstructions of historic climate variables across the United States by spatially-interpolating weather station data \citep{diLuzio2008constructing}. 
We calculated 30-year climate normals for annual and seasonal mean temperature and cumulative precipitation for the recent and historic periods.
\tom{We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December). }{I think this needs more biological justification. I can imagine a simpler divion of cool season and warm season.}
\tom{We additionally calculated the coefficient of variation of each climate driver during each period.}{why? and how? was this based on monthly means or sums?}
We then took the difference between recent and historic periods for the mean and coefficient of variation for each climate driver (Fig. A5).
Because initial analyses indicated a high degree of collinearity between seasonal and annual changes in temperature, we used annual temperature only, along with annual and seasonal precipitation, in the subsequent analysis.

We then evaluated how predicted changes in endophyte prevalence correlated with changes in the environment at an evenly spaced grid of points across the study area.\tom{}{Map?} 
Because regions varied in their predicted starting endophyte prevalence, we calculated the \tom{relative change}{calculated how?} in endophyte prevalence between 1925 and 2020. 
This time period connects the predicted change in prevalence with the endpoints of the available climate record.
We then calculated the Pearson correlation coefficient between the relative change in endophyte prevalence and each climate driver.\tom{}{This paragraph should be clearer that you are using predicted prevalence from the model.}

@@

\section*{Results}

\subsection*{Temporal and spatial trends}

We found that endophyte prevalence increased within the examined specimens over the last two centuries for all three host species (Fig. 2). 
On average, \emph{A. hyemalis} and \emph{E. virginicus} both increased from ~30 \% to over 50\% prevalence across the study region, and \emph{A. perennans} increased from ~ 15\% to over 70\% prevalence.
Rerunning the analysis excluding specimens collected before 1900 when samples are much sparser (Fig. 2A), lead to qualitatively similar predictions, and so we continued analyses with the full dataset. \tom{}{I really like fig 2 but if you are going to split the predictions by latitude you should explain why, talk about this in the results, and also split the points by high and low latitude. Were 35 and 43 the bounds of the data?}

>>>>

	\section*{Abstract}
Many species' distributions and abundances have shifted in response to climate change. 
Most species harbor microbial symbionts that have the potential to influence these responses.
Mutualistic microbial symbionts may provide resilience to environmental change by protecting their hosts from increasing stress. Alternatively, environmental change that causes declines in host or symbiont may disrupt the interaction. 
Microbes can be preserved within the dried plant tissue of herbarium specimens presenting an opportunity to quantify changes across broad temporal and spatial scales. 
We ask how the prevalence of a symbiont of grasses, \emph{Epichloë} fungal endophytes, which can protect hosts from drought, has changed over time in response to climate change, and how these changes vary across the hosts' ranges.
Specifically, we took seed samples from herbarium specimens of three grass host species collected over the last two centuries, quantified fungal hyphae presence within the seeds using microscopy, and evaluated spatial and temporal trends. 
Overall, endophytes have increased in prevalence over the last two centuries from 25\% prevalence to 75\% prevalence on average across the three hosts.
We also found that that the changes in prevalence were associated with observed changes in annual and seasonal climate drivers corresponding with each hosts' peak growing season. 
Thus we provide novel evidence for a cryptic biological response to climate change that may contribute to the resilience of the symbiosis.

@@

	% The journal does not have numbered sections in the main portion of
	% articles. Please refrain from using section references (à la
	% section~\ref{section:CountingOwlEggs}), and refer to sections by name
	% (e.g. section ``Counting Owl Eggs'').

Understanding how biotic interactions are altered by global climate change is a major goal of basic ecological research and conservation management \cite{gilman2010framework,blois2013climate}.
Documented responses to environmental change, such as shifts in species' distributions \cite{aitken2008adaptation} and phenology \cite{piao2019plant}, are typically blind to concurrent changes in associated biotic interactions.
Empirically evaluating these biotic changes, whether interacting species shift in tandem with their partners or not \cite{hillerislambers2013will}, is crucial to predicting future climate change responses, but has been limited because, simply put, data on species interactions are limited, with few datasets extending over the long time scales of contemporary climate change \cite{poisot2021global}.

Natural history collections, which were originally collected to study and preserve taxonomic diversity, present a unique opportunity to explore long-term changes in ecological interactions across broad spatial scales \citep{meineke2018unrealized}. 
Natural history collections, built and maintained by the efforts of thousands of scientists, are invaluable time machines, primarily comprised of physical specimens of organisms  along with information about the time and place of their collection. 
These specimens are samples preserving physical legacies of ecological processes and species' interactions from dynamically changing environments across time and space.
Researchers have used plant collections to document climate change responses, including shifts in phenology \citep{willis2017old, park2019herbarium,  berg2019examination}, as well as in rates of pollination \citep{pauw2011reconstruction, duan2019century}, and herbivory \citep{meineke2019herbarium}. However, focus has been lacking on long-term changes in a particularly common type of biotic interaction, microbial symbioses.

Microbial symbionts are common to all macroscopic organisms and can have important effects on their hosts' survival, growth and reproduction \cite{rodriguez2009fungal,mcfall2013animals}.
Many microbial symbionts act as mutualists, engaging in reciprocally beneficial interactions with their hosts that ameliorate environmental stress. 
For example, bacterial symbionts of insects, such as \emph{Wolbachia}, can improve their hosts' thermal tolerance \citep{truitt2019wolbachia, renoz2019evolutionary}, and arbuscular mycorrhizal fungi, documented in 70-90\% of families of land plants \citep{parniske2008arbuscular}, allow their hosts to persist through drought conditions by improving water and nutrient uptake \citep{cheng2021elucidating}.
On the other hand, changes in the mean and variance of environmental conditions may disrupt microbial  mutualisms by changing the costs and benefits of the interaction for each partner, leading the interaction to deteriorate \citep{aslan2013mutualism}. 
Coral bleaching (the loss of symbiotic algae) due to temperature stress \citep{sully2019global} is perhaps the best known example, but this phenomenon is not unique to corals.
Lichens exposed to elevated temperatures experienced loss of photosynthetic function along with changes in the community composition of their algal symbionts \citep{meyer2022climate}.
How commonly and under what conditions microbial mutualisms deteriorate or strengthen under climate change remain unanswered questions, but previous work suggests that these alternative responses may depend on the intimacy and specialization of the interaction as well as the physiological tolerances of the mutualists \citep{toby2010mutualisms, warren2014mutualism, rafferty2015phenological}.

Understanding of how microbial symbioses are affected by climate change is additionally complicated by spatial hetereogeneity in the direction and magnitude of environmental change \cite{ipcc_2021}. 
Beneficial symbionts are likely able to shield their hosts from environmental stress in locations that experience a small degree of change, but symbionts in locations that experience changes of large magnitude may be pushed beyond their physiological limits \cite{webster2008temperature}.
Additionally, symbionts are often unevenly distributed across their hosts' distribution.
Facultative symbionts may be absent from portions of the host range \cite{afkhami2014mutualist}, and hosts may engage with a diversity of partners (different species, or different locally-adapted strains of one species) across their environments \cite{frade2008variation, rolshausen2018quantifying}.
Identifying broader spatial trends in symbiont prevalence is therefore an important step in developing predictions of where to expect changes in the symbiosis in future climates.

\emph{Epichloë} fungal endophytes are specialized symbionts of cool-season grasses, which have been documented in $\sim 30$\% of cool-season grasses \citep{leuchtmann1992systematics}
There are transmitted vertically from maternal plants to offspring through the grasses' seeds.
Vertical transmission creates a feedback between the fitness of host and symbiont \citep{fine1975vectors, douglas1998host, rudgers2009fungus}. 
Over time, endophytes that act as mutualists for their hosts should rise in prevalence within a population. 
\emph{Epichloë} are known to improve their hosts' drought tolerance \cite{decunta2021systematic}, and protect their hosts against herbivores \cite{crawford2010fungal} and pathogens \cite{xia2018role} likely through the production of a diverse suite of alkaloids and other secondary metabolites.
The fitness feedback induced by vertical transmission leads to the prediction that endophyte prevalence should be high in populations where these fitness benefits are most important. 
Previous survey studies have documented large-scale spatial patterns in  endophyte prevalence structured by environmental gradients \citep{granath2007variation,bazely2007broad, afkhami2012fungal,sneck2017variation}.
We predicted that prevalence should also track temporal changes in environmental drivers that elicit these fitness benefits.
For example, endophyte-mediated drought tolerance should lead prevalence to increase in regions where precipitation declines over time.
Early research on \emph{Epichloë} used herbarium specimens to describe the broad taxonomic diversity of hosts \citep{white1985endophyte}. 
Grasses are commonly identified and collected based on the presence of their reproductive structures meaning that collected specimens typically contain seeds, conveniently preserving the fungi along with their host plants on herbarium sheets. 
This creates the opportunity to leverage the unique spatio-temporal sampling of herbarium collections to examine the response of the symbiosis to historical climate change.

In this study, we assessed for the first time the long-term responses of endophyte symbiosis to climate change through the use of herbarium specimens of three \emph{Epichloë} endophyte host species, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}.
We first address questions describing spatial and temporal trends in endophyte prevalence: (i) How does endophyte prevalence vary across space?, (ii) How has endophyte prevalence changed over time? and (iii) How have temporal changes in endophyte prevalence differed across space?
We evaluated the predictive ability of this climate-implicit model using contemporary endophyte surveys as "out-of-sample test data, an important but rarely used strategy in ecological studies \cite{tredennick2021practical}. 
We then address how climate change may be driving these trends in endophyte prevalence by addressing the question: (iv) What is the relationship between variation in temporal trends in endophyte prevalence and changes in climate drivers?
We examined 1,865 specimens collected across across eastern North America between 1824 and 2019.

\section*{Methods}
        \subsection*{Focal species}
Our surveys focused on three endophyte hosts, \emph{Agrostis hyemalis}, \emph{Agrostis perennans}, and \emph{Elymus virginicus}. Both \emph{Agrostis} species host \emph{Epichloë amarillans} \cite{craven2001multigene, leuchtmann2014nomenclatural}, while \emph{Elymus virginicus} typically hosts \emph{Epichloë elymi} \cite{clay2002evolutionary}.
These $C_3$ grass species are commonly represented in natural history collections with broad distributions covering much the Eastern United States.
\emph{A. hyemalis} is a small short-lived perennial species that germinates in the spring and typically blooms between March and July (most common collection month: May).
\emph{A. perennans} is of similar stature but is longer lived than \emph{Agrostis hyemalis} and blooms in late Summer and early Autumn (most common collection month: September). 
This species is relatively more sparsely distributed, tending to be found in shadier and more moist habitats while \emph{A. hyemalis} is often found in open and recently disturbed ground. 
Both \emph{Agrostis} species are recorded from throughout the Eastern US, but \emph{A. perennans} has a slighty more northern distribution, whereas \emph{A. hyemalis} is found rarely as far north as Canada and is listed as a rare plant in Minnesota.
\emph{E. virginicus} is a larger and relatively longer-lived  species that is more broadly distributed that the \emph{Agrostis} species. 
It begins flowering as early as March or April but continues throughout the Summer (most common collection month: July).

		\subsection*{Herbarium surveys}
We visited herbaria between 2019 and 2022, including The Botanical Research Institude of Texas (which also houses the Vanderbilt University herbarium and the R. Dale Thomas Plant collection from the University of Louisiana at Monroe), the Louisiana State University Herbarium, the Mercer Botanic Garden Herbarium, Missouri Botanic Garden's Herbarium, the S.M. Tracy Herbarium at Texas A\&M, the University of Kansas Herbarium, the Robert Bebb Herbarium at the University of Oklahoma, the Lundell Herbarium and the UT Herbarium at the University of Texas, and Oklahoma State University Herbarium (See Table A1 for a summary of specimens includes from each collection).
Our sampling plan was designed to minimize damage to these specimens.
We chose our focal species in part because they are commonly represented in herbarium collections, and produce high numbers of seeds, meaning that small samples would not diminish the value of the specimens for future studies. 
We collected 5-10 seeds per specimen after examining the herbarium sheet under a dissecting microscope to ensure that we collected viable seeds fit for our purpose of identifying fungal endophytes.

After collecting seeds, we quantified the presence or absence of \emph{Epichloë} fungal hyphae, which grow intercellularly, using microscopy. 
We first softened seeds with a 10\% NaOH solution, then stained the seeds with aniline blue dye and squashed them under a microscope cover slip. 
We examined the squashed seeds for the presence of fungal hyphae at 100X magnification \cite{bacon2018stains}.
In some cases, the tissues examined during microscopy came from flowers or otherwise non-viable seeds, which were excluded from the seed counts for that specimen.
On average we scored 4.7 seeds per specimen in \emph{A. hyemalis}, 5 seeds per specimen in \emph{A. perennans}, and 3.8 seeds per specimen in \emph{E. virginicus}.
Due to imperfect vertical transmission \cite{afkhami2008symbiosis}, it is possible that symbiotic host-plants produce a mixture of symbiotic and non-symbiotic seeds. 
We therefore designated each specimen as symbiotic if endophyte hyphae were observed in one or more seeds, or non-symbiotic if hyphae were observed in zero seeds. 
To capture uncertainty in the identification process, we recorded both a "liberal" and a "conservative" endophyte status for each plant.  
When we identified potential endophytes with unusual morphology, low uptake of stain, or a small amount of fungal hyphae across the scored seeds, we recorded a positive liberal status (more likely to be a true endophyte) and a negative conservative status (less likely to be a true endophyte). 
$89$\% of scored plants had matching liberal and conservative endophyte statuses, reflecting high confidence in endophyte identification.
The following analyses presented in the main text used the liberal statuses, but we repeated all analyses with the conservative statuses which yielded qualitatively similar results (Figure A5). 
We excluded specimens for which information about the collection location and date were unavailable.
In total, we quantified endophyte symbiosis for $964$ \emph{A. hyemalis} specimens collected between 1824 and 2019, $299$ \emph{A. perennans} specimens collected between 1863 and 2017, and $602$ \emph{E. virginicus} specimens collected between 1839 and 2017 (Fig. 1, Fig 2A).
Each specimen was assigned geographic coordinates based on collection information recorded on the herbarium sheet using the geocoding functionality of the ggmap R package \citep{kahle2019package}.
Many specimens had digitized collection information readily available, but for those that did not, we transcribed information from pictures of the specimens. 
Collections were geo-referenced to the nearest county centroid, or nearest municipality when that information was available. 
For a few of the oldest specimens, only information at the state level was available, and so we used the state centoid.

\begin{figure}[H]
	\centering
	\includegraphics[width = \linewidth]{collections_map.png}
	\caption{\textbf{Collection locations of herbarium specimens sampled for endophyte presence absence}. Color designates host species ( \emph{A. hyemalis} (green), \emph{A. perennans} (orange), \emph{E. virginicus} (purple))}
\end{figure}

\subsection*{Assessing spatial and temporal changes in endophyte prevalence}
To quantify spatial and temporal trends in endophyte prevalence, we used an approximate Bayesian method, Integrated Nested Laplace Approximation (INLA), to fit a spatio-temporal model of endophyte prevalence.
Compared to MCMC Bayesian methods, INLA provides a computationally more efficient method of ascertaining model posteriors for certain models that can be formulated as latent Gausssian Models.  \cite{rue2009approximate}. 
Many common statistical models, including structured and unstructured mixed-effects models, can be represented as latent Gaussian Models. 
Fitting models with structured spatial effects is possible with MCMC sampling but can require long computation times, making INLA an effective alternative, which has been used to model spatial patterns in flowering phenology \cite{willems2022forest}, habitat overlap between marine predators and prey \cite{sadykova2017bayesian}, the distribution of temperate trees \cite{engel2022spatial} as well as the population dynamics of endangered amphibians \cite{knapp2016large} and other ecological processes \cite{beguin2012hierarchical}.

We modeled endophyte presence/absence ($P$) as a Bernoulli response variable for specimen $i$ of host species $h$.

@@

The expected endophyte prevalence, $\hat{P}$, was modelled with intercepts specific to each host species ($\beta_{0}$), slopes for changes over time ($\beta_{1}$) as well as the interaction between time and the specimen's latitude and longitude ($\beta_{2}$, $\beta_{3}$, and $\beta_{4}$). 
To quantify spatial patterns in prevalence, we included a spatially-dependent random intercept, $\phi$.
This random spatial effect was modeled as a stochastic partial differential equation (SPDE) that depended on a covariance matrix according to the proximity of each collection location \citep{lindgren2011explicit,bakka2018spatial}. 
The covariance matrix is approximated using a Matérn covariance function, with each data point assigned a location according to the nodes of a mesh of non-overlapping triangles across our study area (Fig A2).
We accounted for potential biases introduced during the process of collecting specimens as well as in scoring ability by including random effects specific to each collector $\chi$ and to each scorer $\omega$.
We performed modeling fitting using the rINLA package, \citep{lindgren2015bayesian}, with vague priors, and compared models with different sizes of mesh, which had little effect on the resulting model estimates.
Posterior modes were stable indicating that numeric convergence was successful.
We assessed model fit with graphical posterior predictive checks (Fig. A3).
The model performed adequately at classifying the historical data (AUC = 0.77; Fig. A4).

We used data from contemporary surveys of on endophyte prevalence  in \emph{A. hyemalis} and \emph{E. virginicus} as test data to evaluate predictions of the model. 
During these surveys, which took place between 2013 and 2020, we collected seeds from living plants from 63 populations(average number of plants sampled: 22.9) across a portion of each species' range (Fig A4). 
We quantified the endophyte status with staining microscopy as described for the herbarium surveys and calculated prevalence of endophytes within the population (proportion of symbiotic plants divided by the number of sampled plants).
We compared the model's prediction for these locations to the observed population prevalence.

		\subsection*{Assessing the role of climate drivers}

We assessed how the magnitude of climate change may have driven changes in endophyte prevalence by assessing correlations between changes in climate and changes in endophyte prevalence predicted from our spatial model at evenly spaced pixels across the study area. 
We first downloaded monthly temperature and precipitation rasters from the PRISM climate group \citep{daly2013prism} covering the time period between 1895 and 2020 using the 'prism' R package \citep{Rprism2015}. 
Prism provides reconstructions of historic climate variables across the United States by spatially-interpolating weather station data \citep{diLuzio2008constructing}. 
We calculated 30-year climate normals for annual and seasonal mean temperature and cumulative precipitation for the recent (1990 to 2020) and historic (1895 to 1925) periods.
We used three four-month seasons within the year (Spring: January, February, March, April; Summer: May; June, July, August; Autumn: September, October, November, December). 
This division of seasons allowed us to quantify differences in climate associated with the two "cool" seasons  that shoulder summer when we expect our focal species to be most biologically active (\emph{A. perennans}: Spring; \emph{E. virginicus}: Spring and Summer; \emph{A. perennans}: Fall). 
In addition to mean climate conditions, environmental variability  in and of itself can influence population dynamics \cite{tuljapurkar_population_1982} and changes in variability are a key prediction of climate change models \cite{stocker2013technical, ipcc_2021}.
So we calculated the coefficient of variation during each period for each annual and seasonal climate driver as the interannual standard deviation divided by the mean across each 30-year period.
We then took the difference between recent and historic periods for the mean and coefficient of variation for each climate driver (Fig. A5).
Because initial analyses indicated a high degree of collinearity between seasonal and annual changes in temperature, we used annual temperature only, along with annual and seasonal precipitation, in the subsequent analysis.
All together, this left us with measurements of change in 10 potential climate drivers: the mean and coefficient of variation of annual temperature, as well as the mean and coefficient of variation of cumulative annual precipitation, cumulative spring precipitation, cumulative summer precipitation, and cumulative autumn precipitation (Fig A8-A9).

We calculated the relative change in endophyte prevalence across the 1925 to 2020, i.e., the difference between predicted prevalence divided by the endophyte prevalence in 1925, because regions varied in their predicted starting prevalence. 
This time period connects the predicted change in prevalence with the endpoints of the available climate record.
We then calculated the Pearson correlation coefficient between the relative change in endophyte prevalence and each climate driver.

@@

\section*{Results}
\subsection*{Temporal and spatial trends}
We quantified temporal and spatial trends in endophyte prevalence while accounting for potential biases introduced by collectors and by individuals who quantified endophyte presence/absence with the use of random effects. We found no evidence that collector biases influenced our results (random effects values centered around 0; Fig 2A), while the identity of individual scorers did contribute to observed patterns in endophyte prevalence (3 of the 16 scorers were more likely than average to assign positive endophyte status, as indicated by 95\% credible intervals that do not overlap 0) (Fig 2B). Interpretation of the data without accounting for this source of variation would lead to potentially biased interpretations of temporal and spatial trends.

\begin{figure}[H]
	\label{fig:random_fx}
	\centering
	\includegraphics[width = .8\linewidth]{random_fx_plot.png}
	\caption{\textbf{Posterior estimates of (A) collector and (B) scorer random effects.} Points show tthe posterior mean along with 95\% CI for random effects estimate from 532 collectors and 16 scorers.}
\end{figure}
